---
title: "Personal-Key-Indicators-of-Heart-Disease"
author: "Buning(Erica) Fan, Jun Ren, Anqi Luo, Jinghan(Tracy) Cui"
date: "2023-04-12"
output: github_document
---
# Abstract
The purpose of this study is to investigate the relationship between heart disease and various factors. Specifically, our research divided our investigation into three aspects: age and sex, BMI, and race. For each aspect, our research analyzed categories including people’s behaviors in their daily lives, their pre-existing medical conditions, and self-ratings to explore their relationship with heart disease. In the Age and Sex aspect, we divided the sample into eight groups based on age and gender, and within each group, we scored their behaviors and diseases to visualize the relationship. In both the BMI (body mass index) and Race aspects, we explored the potential relationship with heart disease. Our project used data visualization techniques throughout our study to effectively present our findings. To avoid subjectivity by only relying on the score, the research included a logistic regression model at the end to better summarize our findings.

# Introduction
Heart disease is a major cause of morbidity and mortality worldwide, and its incidence has been increasing rapidly in recent years. Various factors such as age, gender, lifestyle, and pre-existing medical conditions can significantly impact the development of heart disease. Understanding the relationship between these factors is crucial for identifying effective prevention and treatment strategies. 

In this project, our team wants to investigate how different variables (BMI, race, smoking, alcohol drinking, physical activity, Asthma, Diabetic, Difficulty of Walking, Kidney Disease, Skin Cancer, and Stroke, physical health, mental health, sex, and age group) affect the probability of having heart diseases. To achieve this, we will use the Key Indicators of Heart Disease dataset from the CDC, which consists of survey data from 400k adults related to their health status.

Our main questions include: How do behavioral habits and pre-existing medical conditions impact the incidence of heart disease? How does age and gender affect the relationship between behavioral habits, pre-existing medical conditions, and heart disease incidence? What is the relationship between BMI and heart disease incidence, and how does this vary across age, gender, and race? How do different races and ethnicities vary in their incidence of heart disease and their risk factors? What recommendations can we provide for patients on how to prevent heart disease based on these key indicators? 

# Data & Methods
## Data
- The dataset comes from the CDC and is a major part of the Behavioral Risk Factor Surveillance System (BRFSS), which conducts annual telephone surveys to gather data on the health status of U.S. residents. As the CDC describes: "Established in 1984 with 15 states, BRFSS now collects data in all 50 states as well as the District of Columbia and three U.S. territories. BRFSS completes more than 400,000 adult interviews each year, making it the largest continuously conducted health survey system in the world.". The most recent dataset (as of February 15, 2022) includes data from 2020. It consists of 401,958 rows and 279 columns. 

- URL: https://www.kaggle.com/datasets/kamilpytlak/personal-key-indicators-of-heart-disease
Personal Key Indicators of Heart Disease | Kaggle

- Variables:
    - HeartDisease - target trait.

    - Sex - gender of a person.

    - AgeCategory - age category of the subjects.

    - BMI - a value derived from the mass and height of a person.

    - Race - imputed race/ethnicity value.

    - Behaviors:
      - Smoking - whether the participant has smoking.

      - AlcoholDrinking - whether the participant has alcohol drinking.

      - Physical Activity - participants who have physical activity in the past month.

    - Other Disease:

      - Stroke - Ischemic stroke occurs 4 times more often than hemorrhagic. 

      - DiffWalking - difficulty climbing stairs.

      - Diabetic - whether the participant has diabetes before/now.

      - Asthma - whether the participant has asthma before/now.

      - KidneyDisease - Whether the participant had kidney disease before/now (Not including kidney stones, bladder infection or incontinence).

      - Skin Cancer - whether the participant has skin cancer before/now.

    - PhysicalHealth - number of days in a month of feeling poor physical health.

    - MentalHealth - number of days in a month of feeling poor mental health.

    - GenHealth - individual general health rate.

    - SleepTime - average number of  hours of sleep in a 24-hour period.
    
## Method
Overall, the data is summarized into three big groups: AgeCategory/Sex, BMI, and Race. We use these three different aspects to find out the relationship between each of those three aspects and heart disease.

### In Age Category & Sex aspect:
We are currently categorizing various factors into four groups: behavioral habits, diseases, self-assessment of physical and mental health, and BMI. To assess the overall effect, we are scoring the behavioral habits and diseases separately before adding the scores. This approach allows us to differentiate the contributions of lifestyle choices, pre-existing medical conditions, and individuals’ perceptions of their health to the outcome. However, since self-assessment is based on individuals' subjective experiences of their physical and mental health issues within a 30-day period, we have decided not to include self-assessment scores in the same manner as the other two groups for our analysis. This is because these scores may be less objective, as personal perceptions can vary and be influenced by multiple factors. Similarly, we are considering BMI based on the individual's situation related to their mass and height. The categorization of those key indicators are: 

1. Behavior: Smoking; Alcohol Drinking; Physical Activity
2. Disease: Stroke; Diff Walking; Diabetic; Asthma; Kidney Disease; Skin Cancer
3. Self-assessment: Physical Health; Mental Health
4. BMI

#### Evaluation of Behaviors & Diseases  
For each age and gender category, the data are organized by behavioral and morbidity classifications. We intend to evaluate both behavioral and morbidity classifications. Within each classification, we calculate and depict the proportion of individuals who have or have not engaged in any specific behavior or experienced any disease using graphs. Behaviors/diseases are categorized as positive or negative. Each behavior/disease is assigned a proportional value, ranging from 0 to 1. The evaluation for each category results from the sum of its contained behaviors/diseases.

- Since the range of proportional values for behaviors is from 0 to 0.6, we have divided the proportional value range into six equal intervals for Smoking, AlcoholDrinking, and PhysicalActivity. Each interval corresponds to a point on the scale:

```{r echo=FALSE}
table1 <- data.frame(
  Proportional_Value = c("0 - 0.1", "0.1 - 0.2", "0.2 - 0.3", "0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6"),
  Corresponding_Score = c(6, 5, 4, 3, 2, 1)
)
print(table1)
```

- The proportional values for diseases (Stroke, DiffWalking, Diabetic, Asthma, KidneyDisease, Skin Cancer) are set into six values ranging from 0 to 0.3. As these values are all less than 0.4, each corresponding point is set below this threshold:

```{r echo=FALSE}
table2 <- data.frame(
  Proportional_Value = c("0 - 0.05", "0.05 - 0.1", "0.1 - 0.15", "0.15 - 0.2", "0.2 - 0.25", "0.25 - 0.3"),
  Corresponding_Score = c(6, 5, 4, 3, 2, 1)
)
print(table2)
```

We evaluated each classification and recorded the evaluation score for each behavior and disease. After obtaining the scores for these two categories, we compared the relationship between the behavior/disease scores and the age category. Once we identified the trend of the scores, we then compared the relationship between the proportion of heart disease and age category to observe their correlation. Additionally, we also analyzed the relationship between self-assessment and BMI with AgeCategory/Sex to determine their respective associations.

### In BMI aspect:
In the second part of our study, we aimed to investigate the relationship between BMI and other aspects. In the 1990s, the World Health Organization (WHO) introduced the concept of BMI, which defines a BMI less than 18.5 as underweight, 18 to 24 as healthy weight, 25 to 30 as overweight, and over 30 as obese. Using this definition, we divided BMI values into several groups: 10-18; 18-24; 24-29; 29-34; 34-39; 39-45; 45-50; 50-60; and 60 to infinity.

Following the same category in Age Category & Sex aspect, we keep using the same categories of behaviors and other diseases and compare those two proportional values with BMI values. We calculate the mean of each proportional classifications in behaviors(Smoking, PhysicalActivity, AlcoholDrinking) and diseases(HeartDisease, Stroke, Diffwalking, Diabetic, Asthma, KidneyDisease, SkinCancer). We observe the relationship between BMI and Behaviors and Diseases by line plots. We also particularly check the relationship between BMI and heart disease by dot plot.

### In Race aspect:
The third part aims to investigate the relationship between race and heart disease, which is an interesting fact. The race aspect is divided into six categories: American Indian/Alaskan Native, Asian, Black, Hispanic, White, and Other. We imported the data of the number of people who have heart disease in each category and analyzed if race has any influence on heart disease. We used dot plots and covariance graphs for this analysis.

## Data importing and cleaning
### Set Up and Import Data
```{r include=FALSE, echo=FALSE}
library(tibble)
library(tidyr)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(knitr)
data <- read.csv("data/heart_2020_cleaned.csv")
```

### Filter out NA Data
```{r include=FALSE, echo=FALSE}
str(data)
summary(data)
data[rowSums((is.na(data))>0),]
na<-function(x){
  result<- which(x=='na'| x=='NA'|x=='Na')
  length(result)
}
apply(data,2,na)
```

### Change Yes/No Response to 1/0
```{r echo=FALSE}
# creating dummy variables using a function
# change Yes to 1, No to 0
convert_yes_no_to_binary <- function(data, column) {
  for (x in column) {
    data[[x]] <- ifelse(data[[x]] == 'Yes', 1, 0)
  }
  return(data)
}

dummy_variables <- c("HeartDisease", "Smoking", "AlcoholDrinking", "Stroke", "DiffWalking","Diabetic", "Asthma", "KidneyDisease", "SkinCancer")

# physical activity has a relative positive impact, so change Yes to 0, No to 1
data$PhysicalActivity <- ifelse(data$PhysicalActivity == "Yes", 0, 1)

data <- convert_yes_no_to_binary(data, dummy_variables)
head(data)
```

### Seperate Into 8 Different Group Based on Age and Sex
```{r include=FALSE, echo=FALSE}
# check unique values in AgeCategory and Sex
unique(data$AgeCategory)
length(unique(data$AgeCategory))
unique(data$Sex)

# update AgeCategory variable
data <- data %>%
  mutate(AgeCategory = case_when(
    AgeCategory %in% c("18-24", "25-29", "30-34", "35-39") ~ "under 40",
    AgeCategory %in% c("40-44", "45-49", "50-54") ~ "40-54",
    AgeCategory %in% c("55-59", "60-64", "65-69") ~ "55-69",
    AgeCategory %in% c("70-74", "75-79", "80 or older") ~ "over 70",
    TRUE ~ "Unknown"))

# create Combined_AgeSexCategory variable using AgeCategory and Sex
data <- data %>%
  mutate(Combined_AgeSexCategory = case_when(
    AgeCategory == "under 40" & Sex == "Female" ~ "Female under 40",
    AgeCategory == "under 40" & Sex == "Male" ~ "Male under 40",
    AgeCategory == "40-54" & Sex == "Female" ~ "Female 40-54",
    AgeCategory == "40-54" & Sex == "Male" ~ "Male 40-54",
    AgeCategory == "55-69" & Sex == "Female" ~ "Female 55-69",
    AgeCategory == "55-69" & Sex == "Male" ~ "Male 55-69",
    AgeCategory == "over 70" & Sex == "Female" ~ "Female over 70",
    AgeCategory == "over 70" & Sex == "Male" ~ "Male over 70",
    TRUE ~ "Unknown"))

head(data)
```

### Create 8 Group Tables
```{r echo=FALSE}
female_under_40 <- data %>%
  filter(Combined_AgeSexCategory == "Female under 40")

male_under_40 <- data %>%
  filter(Combined_AgeSexCategory == "Male under 40")

female_40_54 <- data %>%
  filter(Combined_AgeSexCategory == "Female 40-54")

male_40_54  <- data %>%
  filter(Combined_AgeSexCategory == "Male 40-54")

female_55_69 <- data %>%
  filter(Combined_AgeSexCategory == "Female 55-69")

male_55_69 <- data %>%
  filter(Combined_AgeSexCategory == "Male 55-69")

female_over_70 <- data %>%
  filter(Combined_AgeSexCategory == "Female over 70")

male_over_70 <- data %>%
  filter(Combined_AgeSexCategory == "Male over 70")
```

# Results
## In Age Category & Sex aspect:
### Female Behaviors
```{r echo=FALSE}
# combine the data frames vertically
female_combined <- rbind(female_under_40, female_40_54, female_55_69, female_over_70)

# calculate proportions of "Yes"
props <- female_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Smoking = mean(Smoking),
    AlcoholDrinking = mean(AlcoholDrinking),
    PhysicalActivity = mean(PhysicalActivity)
  ) %>%
  gather(key = "Behavior", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70"))

# create the bar plot
ggplot(props, aes(x = Behavior, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#FEE5D9", "#FCBBA1", "#FC9272", "#DE2D26"), name = "Age/Sex Category") +
  labs(x = "Behavior", y = "Proportion of Having the Behavior", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Female Participants Who Have Each Behavior")
```

### Male Behaviors
```{r echo=FALSE}
# combine the data frames vertically
male_combined <- rbind(male_under_40, male_40_54, male_55_69, male_over_70)

# calculate proportions of "Yes"
props <- male_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Smoking = mean(Smoking),
    AlcoholDrinking = mean(AlcoholDrinking),
    PhysicalActivity = mean(PhysicalActivity)
  ) %>%
  gather(key = "Behavior", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Male under 40", "Male 40-54", "Male 55-69", "Male over 70"))

# create the bar plot
ggplot(props, aes(x = Behavior, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#EFF3FF", "#BDD7E7", "#6BAED6", "#3182BD"), name = "Age/Sex Category") +
  labs(x = "Behavior", y = "Proportion of Having the Behavior", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Male Participants Who Have Each Behavior")
```

The proportional values of the behavioral category are presented by bar graphs in the female and male categories. The red trend represents the female behavioral category, and the blue trend represents the male behavioral category. Throughout both bar graphs, the overall trend between females and males is similar. In the female graph, the proportion of physical activity for females has an increasing trend. The proportion of smoking for females also has an increasing trend at first, but it decreases when the age of females is over 70. In the male graph, both proportions of physical activity and smoking continue to increase as they get older.

### Female Disease
```{r echo=FALSE}
# calculate proportions of "Yes"
props <- female_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Stroke = mean(Stroke),
    DiffWalking = mean(DiffWalking),
    Diabetic = mean(Diabetic),
    Asthma = mean(Asthma),
    KidneyDisease = mean(KidneyDisease),
    SkinCancer = mean(SkinCancer)
  ) %>%
  gather(key = "Disease", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70"))

# create the bar plot
ggplot(props, aes(x = Disease, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#FEE5D9", "#FCBBA1", "#FC9272", "#DE2D26"), name = "Age/Sex Category") +
  labs(x = "Disease", y = "Proportion of Having the Disease", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Female Participants Who Responded 'Yes' to Each Disease")
```

### Male Disease
```{r echo=FALSE}
# calculate proportions of "Yes"
props <- male_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Stroke = mean(Stroke),
    DiffWalking = mean(DiffWalking),
    Diabetic = mean(Diabetic),
    Asthma = mean(Asthma),
    KidneyDisease = mean(KidneyDisease),
    SkinCancer = mean(SkinCancer)
  ) %>%
  gather(key = "Disease", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Male under 40", "Male 40-54", "Male 55-69", "Male over 70"))

# create the bar plot
ggplot(props, aes(x = Disease, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#EFF3FF", "#BDD7E7", "#6BAED6", "#3182BD"), name = "Age/Sex Category") +
  labs(x = "Disease", y = "Proportion of Having this Disease", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Male Participants Who Responded 'Yes' to Each Disease")
```

The two bar graphs represent the proportion of both female and male participants who have particular diseases. Same with the behavior graphs: the red trend represents the female, and the blue trend represents the male. Both graphs are similar in their tendency. The proportion of participants with asthma in both females and males presents a decreasing tendency as they get older. All of the other diseases (Diabetes, Difficulty walking, Kidney Disease, Skin Cancer, Stroke) show an upward trend as the participants get older.

Based on the evaluation rubric presented in the methods section, we have calculated the evaluation scores for each behavioral and disease category by Age/Sex category. The evaluation scores are obtained by summing up the corresponding points for each category. The detailed evaluation results are presented in the following table.

### Evaluate Each Group and Collect Score
```{r warning=FALSE, message=FALSE, echo=FALSE}
score <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Behavior_Score = c(15, 13, 12, 12, 14, 13, 12, 11),
  Disease_Score = c(33, 30, 26, 21, 34, 32, 26, 19),
  Total_Score = c(48, 43, 38, 33, 48, 45, 38, 30)
)

score

score_sep <- score %>%
  separate(Group, c("Gender", "Age"), sep = " ", remove = FALSE)
score_sep$Age <- recode(score_sep$Age, "under" = "under 40")
score_sep$Age <- recode(score_sep$Age, "over" = "over 70")
score_sep$Age <- factor(score_sep$Age, levels = c("under 40", "40-54", "55-69", "over 70"))

line_data <- score_sep %>%
  select(-Group) %>%
  tidyr::gather(key = "Score_Type", value = "Score", -Gender, -Age)

# create the line plot
ggplot(line_data, aes(x = Age, y = Score, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Score") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  scale_color_manual(values = c("Female" = "magenta", "Male" = "blue"), name = "Gender") +
  facet_wrap(~Score_Type, ncol = 1) +
  ggtitle("Evaluation Score by Age Group")
```

The total scores indicate that younger age groups have the highest scores, whereas older age groups have the lowest scores. Furthermore, both male and female groups over 70 years of age have the lowest scores compared to their counterparts in other age groups.

We created three line graphs, including the behavior score, disease score, and total score by Age/Sex Category to illustrate the trends in the scores. The graphs indicate that the behavior score for both females and males across different Age/Sex categories only slightly decreases. The disease score, however, exhibits an apparent decline as the age of both females and males increases, as shown by the total score graph. Additionally, there is no significant difference between the male and female lines in the three line plots.

### Physical Health Rating
```{r echo=FALSE}
# calculate means of PhysicalHealth by age/sex category
means <- data %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    PhysicalHealth_mean = mean(PhysicalHealth)
  )

age_groups <- c("under 40", "40-54", "55-69", "over 70")

colors <- c("Female" = "#DE2D26", "Male" = "#3182BD")

# create the line plot
mean_data <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Mean = means$PhysicalHealth_mean[c(4, 1:3, 8, 5:7)],
  Gender = c("Female", "Female", "Female", "Female", "Male", "Male", "Male", "Male"),
  Age = age_groups
)

ggplot(mean_data, aes(x = Age, y = Mean, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Average Physical Health Rating") +
  theme_bw() +
  scale_color_manual(values = colors, name = "Gender") +
  scale_x_discrete(limits = age_groups) +
  ggtitle("Average of Physical Health Rating for Female and Male on Age Range")
```

### Mental Health Rating
```{r echo=FALSE}
# calculate means of MentalHealth by age/sex category
means <- data %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    MentalHealth_mean = mean(MentalHealth)
  )

age_groups <- c("under 40", "40-54", "55-69", "over 70")

colors <- c("Female" = "#DE2D26", "Male" = "#3182BD")

# create the line plot
mean_data <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Mean = means$MentalHealth_mean[c(4, 1:3, 8, 5:7)],
  Gender = c("Female", "Female", "Female", "Female", "Male", "Male", "Male", "Male"),
  Age = age_groups
)

ggplot(mean_data, aes(x = Age, y = Mean, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Average Mental Health Rating") +
  theme_bw() +
  scale_color_manual(values = colors, name = "Gender") +
  scale_x_discrete(limits = age_groups) +
  ggtitle("Average of Mental Health Rating for Female and Male on Age Range")
```

The two line plots depict the trends in individual physical and mental health ratings over a 30-day period. The data presented in the plots represent the average ratings reported by individuals who have reported experiencing physical or mental health problems in a given month, for both males and females. The first plot displays the average rating of physical health, which shows an observable increasing trend for both genders, indicating that individuals feel they have more physical problems as they age. The second plot displays the average rating of mental health, which has an inverse trend as compared to the physical health plot. The trend line shows that individuals feel they have fewer mental health problems as they age. However, the average score for females is consistently higher than that for males, indicating that females tend to feel they have more physical and mental health issues than males.

### Different Group & Sleeptime
```{r echo=FALSE}
data$Combined_AgeSexCategory <- factor(data$Combined_AgeSexCategory, levels = c("Female under 40", "Male under 40", "Female 40-54", "Male 40-54", "Female 55-69", "Male 55-69", "Female over 70", "Male over 70"))

# create the box plot
ggplot(data, aes(x = Combined_AgeSexCategory, y = SleepTime, fill = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", color = "black", fill = "white", position = position_dodge(0.75)) +
  labs(x = "Age Sex Category", y = "SleepTime", fill = "Sex") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  coord_cartesian(expand = FALSE) +
  ggtitle("Boxplot of Sleep Time by Age Sex Category")
```

The box plot above shows sleep time by Age/Sex Category. The plot indicates that there is no significant trend in sleep time, with only a slight increase observed among older age groups for both males and females.

### Different Groups & BMI
```{r echo=FALSE}
data$Combined_AgeSexCategory <- factor(data$Combined_AgeSexCategory, levels = c("Female under 40", "Male under 40", "Female 40-54", "Male 40-54", "Female 55-69", "Male 55-69", "Female over 70", "Male over 70"))

# create the box plot
ggplot(data, aes(x = Combined_AgeSexCategory, y = BMI, fill = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", color = "black", fill = "white", position = position_dodge(0.75)) +
  labs(x = "Age Sex Category", y = "BMI", fill = "Sex") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  coord_cartesian(expand = FALSE) +
  ggtitle("BMI by Age Sex Category")
```

Moreover, the box plot shows the summary of BMI across different age and sex categories. The trend appears to be unclear and random, with no clear pattern or tendency observed.

### Proportion of Getting Heart Disease in Different Age Group
```{r echo=FALSE}
# calculate the proportion of people with heart disease in each group
female_under_40_prop <- sum(female_under_40$HeartDisease == 1) / nrow(female_under_40)
female_40_54_prop <- sum(female_40_54$HeartDisease == 1) / nrow(female_40_54)
female_55_69_prop <- sum(female_55_69$HeartDisease == 1) / nrow(female_55_69)
female_over70_prop <- sum(female_over_70$HeartDisease == 1) / nrow(female_over_70)
male_under_40_prop <- sum(male_under_40$HeartDisease == 1) / nrow(male_under_40)
male_40_54_prop <- sum(male_40_54$HeartDisease == 1) / nrow(male_40_54)
male_55_69_prop <- sum(male_55_69$HeartDisease == 1) / nrow(male_55_69)
male_over70_prop <- sum(male_over_70$HeartDisease == 1) / nrow(male_over_70)

# combine the proportion data
prop_data <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Proportion = c(female_under_40_prop, female_40_54_prop, female_55_69_prop, female_over70_prop,
                 male_under_40_prop, male_40_54_prop, male_55_69_prop, male_over70_prop),
  Gender = c("Female", "Female", "Female", "Female", "Male", "Male", "Male", "Male"),
  Age = c("under 40", "40-54", "55-69", "over 70", "under 40", "40-54", "55-69", "over 70")
)

# create the line plot
ggplot(prop_data, aes(x = Age, y = Proportion, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Proportion of people with heart disease", title= "Proportion of Getting Heart Disease in Different Age Group and Gender") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  scale_color_manual(values = c("Female" = "magenta", "Male" = "blue"), name = "Gender") +
  scale_x_discrete(limits = c("under 40", "40-54", "55-69", "over 70"))
```

The line graph illustrates the relationship between age and the proportion of individuals with heart disease, with separate lines for males and females. The overall pattern indicates an upward trend, suggesting an increase in the risk of developing heart disease with age for both sexes. However, there are notable differences in the rate of increase between males and females. Specifically, the graph reveals that the proportional values of heart disease for both genders are similar for those under 40. Beyond that point, the prevalence of heart disease in males increases at a higher rate, suggesting that males may be more susceptible to the condition as they age.

## In BMI aspect:
### BMI and Behavior
```{r echo=FALSE}
behavior_data <- data %>%
  mutate(BMI_Group = cut(BMI, c(0, 10, 18, 24, 29, 34, 39, 45, 50, 60, Inf))) %>%
  group_by(BMI_Group) %>%
  summarize(HSmoking = mean(Smoking),
            AlcoholDrinking = mean(AlcoholDrinking),
            PhysicalActivity = mean(PhysicalActivity)) %>%
  pivot_longer(-BMI_Group, names_to = "Behavior", values_to = "Proportion")

# create a line plot with different colored lines for each behavior
ggplot(behavior_data, aes(x = BMI_Group, y = Proportion, color = Behavior, group = Behavior)) +
  geom_line() +
  labs(x = "BMI", y = "Proportion of Behavior", color = "Behavior") +
  theme_bw() +
  ggtitle("Proportion of Each Behavior by BMI")
```

The line graph displays the relationship between three behaviors (Alcohol Drinking, Smoking, and No Physical Activity) and different ranges of BMI. The graph reveals that when BMI is in the healthy range of 18-24, people tend to have a lower proportion of smoking and no physical activity. However, as BMI increases from 24 to infinity, the proportion of people who do not engage in physical activity increases, while the trend in proportion of smokers remains relatively unchanged. Additionally, the trend for alcohol consumption shows no significant change across all BMI ranges, suggesting that there may be no strong correlation between alcohol consumption and BMI in this dataset.

Therefore, the graph suggests that as BMI moves into the healthy range, people tend to have a greater level of physical activity and smoke less. As BMI increases, people tend to become less physically active, which may contribute to higher rates of obesity and other health issues. Additionally, the lack of a significant correlation between alcohol consumption and BMI suggests that other factors may also influence smoking.

### BMI and Other Diseases
```{r warning=FALSE, message=FALSE, echo=FALSE}
# create a data frame with BMI groups and proportion of people getting each disease
disease_data <- data %>%
  mutate(BMI_Group = cut(BMI, c(0, 10, 18, 24, 29, 34, 39, 45, 50, 60, Inf))) %>%
  group_by(BMI_Group) %>%
  summarize(HeartDisease = mean(HeartDisease),
            Stroke = mean(Stroke),
            DiffWalking = mean(DiffWalking),
            Diabetic = mean(Diabetic),
            Asthma = mean(Asthma),
            KidneyDisease = mean(KidneyDisease),
            SkinCancer = mean(SkinCancer)) %>%
  pivot_longer(-BMI_Group, names_to = "Disease", values_to = "Proportion")

# create a line plot with different colored lines for each disease
ggplot(disease_data, aes(x = BMI_Group, y = Proportion, color = Disease, group = Disease)) +
  geom_line() +
  labs(x = "BMI", y = "Proportion of Disease", color = "Disease") +
  theme_bw() +
  ggtitle("Proportion of Getting Each Disease by BMI")
```

This line graph shows the proportion of people with different BMI ranges who have been diagnosed with Asthma, Diabetic, Difficulty of Walking, Heart Disease, Kidney Disease, Skin Cancer, and Stroke. The plot reveals interesting insights about the relationship between BMI and diseases. When the BMI is in the range of 18-24, the proportion of people with all diseases, except for Skin Cancer, is lower. This suggests that maintaining a healthy weight in this BMI range could help reduce the risk of developing these diseases. Also, as the BMI increases from 24 to infinity, the proportion of people who have Asthma, Diabetic, Difficulty of Walking, Heart Disease, Kidney Disease, and Stroke increases, with the highest increase observed in Asthma, Diabetic, and Difficulty of Walking. This indicates that being overweight or obese in this range can significantly increase the risk of developing these diseases. However, the only disease that shows a decreasing trend with increasing BMI is Skin Cancer. The reason might because that people with higher BMIs tend to have more subcutaneous fat, which provides natural protection against the harmful effects of UV radiation from the sun.

### BMI and HeartDisease
```{r echo=FALSE, message=FALSE}
# create a new column with BMI group
data <- data %>%
  mutate(BMI_group = cut(BMI, breaks = c(10, 18, 24, 29, 34, 39, 45, 50, 60, Inf)))

# calculate the proportion of people getting HeartDisease
df <- data %>%
  group_by(BMI_group) %>%
  summarize(prop = mean(HeartDisease))

# create the dot plot
ggplot(df, aes(x = BMI_group, y = prop)) +
  geom_line() +
  geom_point() +
  labs(x = "BMI", y = "Proportion of people with HeartDisease") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Proportion of People with Heart Disease by BMI")
```

This plot takes a closer look at the proportion of getting Heart Disease as the BMI grows. The plot indicates that when BMI is in the range of 18-24, the proportion of getting Heart Disease is relatively low, around 0.06. As BMI increases, the proportion of getting Heart Disease also increases, and this relationship appears to be approximately linear. This means that as BMI increases, the risk of getting Heart Disease increases as well.

Overall, these three plots highlight the people with a healthy BMI tend to have a healthier lifestyle and the importance of maintaining a healthy weight in the BMI range of 18-24 to reduce the risk of developing various diseases. It also underscores the need for early intervention and preventative measures for people with BMIs in the overweight and obese range to prevent the onset of chronic diseases.

## In Race aspects:
### Proportion of Getting Heart Disease for Different Races and Age Groups
```{r warning=FALSE, message=FALSE, echo=FALSE}
data_HD <- data %>% 
  mutate(prop_heartdisease = ifelse(HeartDisease == 1, 1, 0)) %>%
  group_by(AgeCategory, Race) %>%
  summarize(prop_heartdisease = mean(prop_heartdisease))

age_levels <- c("under 40", "40-54", "55-69", "over 70")

# create the co-variation plot
data_HD %>% 
  transmute(AgeCategory = factor(AgeCategory, levels = age_levels), Race, 
            prop_heartdisease, fill = prop_heartdisease) %>%
  ggplot(aes(x = AgeCategory, y = Race, fill = fill)) +
  geom_tile() +
  scale_fill_gradient(low = "#FCBBA1", high = "red") +
  labs(x = "Age Category", y = "Race", fill = "Proportion of Heart Disease") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Proportion of People with Heart Disease by Race and Age Category")
```

The plot depicts the covariation between race and the proportion of individuals with heart disease as they age. As expected, the plot shows that the proportion of people getting heart disease increases as age increases, indicated by the darkening color of the plot for all races. However, notable differences are observed among different races. The plot suggests that the Asian race tends to have a lighter color, indicating a lower proportion of people getting heart disease compared to other races. In contrast, the American Indian/Alaskan Native race tends to have a slightly darker color, suggesting a slightly higher proportion of people getting heart disease compared to other races.

### Race and Heart Disease
```{r echo=FALSE}
data %>%
  group_by(Race) %>%
  summarise(prop_heartdisease = mean(HeartDisease)) %>%
  ggplot(aes(x = Race, y = prop_heartdisease, color = Race)) +
  geom_point(size = 3) +
  labs(x = "Race", y = "Proportion of Heart Disease") +
  theme_bw() +theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Proportion of People with Heart Disease by Race")
```

We also draw a dot plot that only shows the relationship between different races and the proportion of getting heart disease. It also appears that there are significant differences between the races with respect to the proportion of getting heart disease. The plot reveals that Asian individuals tend to have a lower proportion of getting heart disease compared to all the other races. This is followed by Hispanic, Black, Other, White, and American Indian/Alaskan Native, respectively. The American Indian/Alaskan Native population has the highest proportion of getting heart disease.

Overall, both plots suggest that race may play a role in the proportion of people getting heart disease as age grows. Further investigation may be needed to understand the underlying factors that contribute to this difference in risk between different races.

## Regression:
### Logistic Regression Model
```{r echo=FALSE}
# fitting the full model
reg_formula <- HeartDisease ~ BMI + SleepTime + Smoking + AlcoholDrinking + PhysicalActivity + DiffWalking + Diabetic + Asthma + KidneyDisease + SkinCancer + Stroke + PhysicalHealth + MentalHealth + AgeCategory + Sex + Race + GenHealth

full_model <- glm(reg_formula, data = data, family = binomial(link = "logit"))
summary(full_model)
```
Since there are two continuous variables (BMI and SleepTime), two discrete variables (PhysicalHealth and MentalHealth), and all other variables being categorical, we fitted a logistic regression model to the data, which we refer to as the full model. By comparing the p-values of each predictor to an alpha value of 0.05, we can conclude that the significant predictors in this model include BMI, SleepTime, Smoking, AlcoholDrinking, DiffWalking, Diabetic, Asthma, KidneyDisease, SkinCancer, Stroke, PhysicalHealth, MentalHealth, AgeCategory, Sex, and Race.

### Variable Selection
```{r echo=FALSE}
# variable selection using stepwise AIC
library(MASS)
stepwise_model <- stepAIC(full_model, direction = "both")
summary(stepwise_model)
```
To exclude all redundant variables in the model and improve its performance, we performed a stepwise variable selection using AIC as a criterion. The stepwise model removed PhysicalHealth, and all other predictors remained.

The deviance residuals provide a measure of the goodness-of-fit of the model. Smaller deviance residuals generally indicate better fit. According to the output, the residual deviance of our model is 146,063, while the null deviance is 186,906. This suggests that the model fits the data better than a model without any predictors.

Also, since smaller values of AIC indicate a better balance between fit and complexity. In the stepwise model, the AIC is 146,115 which is smaller compared to that of the full model.

### Check for multicollinearity using Variance Inflation Factor (VIF)
```{r echo=FALSE}
library(car)
vif_values <- vif(stepwise_model)
vif_values
```
After calculating the Generalized Variance Inflation Factors (GVIF) values in the model, we found that the GVIF^(1/(2*Df)) column provides the adjusted VIF values for each predictor and all values are less than 10, indicating that multicollinearity is not a significant concern in our model.

### Model interpretation
```{r echo=FALSE}
# check odds ratios for each predictor
exp(coef(stepwise_model))
```
The odds ratios for each predictor in the stepwise model indicate how the odds of having heart disease increase or decrease with a one-unit change in the predictor, holding other predictors constant. For example, the coefficient of BMI (0.0055895) is positive, suggesting that an increase in BMI may raise the likelihood of heart disease. Specifically, we used exp(coef(stepwise_model)) to calculate the odds ratios for each predictor. For BMI, we computed exp(0.0055895) = 1.005605133, meaning the odds ratio is approximately 1.005605133. In other words, for every unit increase in BMI, the odds of heart disease increase by roughly 0.56%. Similarly, the odds ratio for SleepTime is 0.9812, indicating that for each additional hour of sleep, the odds of having heart disease decrease by about 1.88%.

By examining the odds ratios, we can determine which predictors have a more significant impact on the probability of developing heart disease. This helps identify which groups of people face higher risks of heart disease and informs strategies to prevent the condition through better daily habits and other methods. For example, compared to people under 40 who have an odds ratio of 0.344007023, those aged over 70 are more likely to develop heart disease, with a much higher odds ratio of 4.458962516. Additionally, alcohol consumption (odds ratio 0.765959330) is not as severe a risk factor as having a stroke, which has the highest odds ratio among all other diseases at 2.900803065. This means that each additional stroke increases the odds of heart disease substantially.

### Perform the Likelihood Ratio Test 
```{r echo=FALSE}
# compare the stepwise model with the full model using ANOVA
anova(stepwise_model, full_model, test = "Chisq")
```
Finally, we performed a likelihood ratio test by comparing the stepwise model with the full model using ANOVA. According to the output, the p-value is 0.9548, which is greater than the alpha value of 0.05, indicating that there is no statistically significant difference between the full model and the stepwise model. Therefore, it can be concluded that the stepwise model, with fewer predictors, provides a similar goodness of fit as the full model. This implies that the stepwise model is more parsimonious and may be preferred over the full model.

# Discussion
Heart disease is a complex and multifactorial condition influenced by a range of factors such as age, gender, family history, smoking, BMI, other diseases, physical inactivity, race, and stress. Our study aims to develop a predictive model to identify the key indicators with the strongest association with heart disease risk and provide recommendations for healthcare providers and patients to prevent heart disease, utilizing the Key Indicators of Heart Disease dataset.

Our goal is to develop a predictive model that can identify the key indicators with the strongest association with heart disease risk. By analyzing the available data, we hope to provide recommendations for healthcare providers and patients to help prevent heart disease. However, it's important to acknowledge the limitations of our dataset.

One limitation of our study was the size of the dataset. While we analyzed a significant amount of data, a larger dataset with more data points could have provided more robust results. Additionally, the distribution of different races in our dataset was highly unbalanced, which might affect the results of the relationship between race and heart disease. Also, the age section only contained age ranges instead of specific ages. Further data collection with detailed ages would enable a more in-depth analysis. Moreover, the dataset contained fewer data points for individuals under 40, which may have introduced bias into our analysis. Having too many predictors in the logistic regression model is also a concern.

Future research should explore the relationship between mental health issues, such as depression, and heart disease. Expanding data collection to include a more diverse population worldwide could provide valuable insights into the multifactorial nature of heart disease. Despite these limitations, our project has the potential to make a significant contribution to our understanding of heart disease risk and prevention. We may also employ feature selection techniques to penalize less important predictors and select the most relevant ones, and use cross-validation to get a better estimate of the model’s performance on unseen data.

To manage heart disease, we recommend people maintain a healthy weight, exercise regularly, quit smoking, get enough sleep, manage diabetes, and limit alcohol intake. Healthcare providers can offer personalized advice to prevent or manage heart disease.

In conclusion, according to the logistic regression model, males are about 2 times more likely to have heart disease compared to females. People with behaviors such as smoking, diseases like stroke, and those aged over 70 with poor general health are more likely to have heart disease. Our study sheds light on the factors associated with heart disease and provides some recommendations for prevention. Further research with a larger and more diverse population is necessary to gain a more comprehensive understanding of heart disease.

# Appendix
## Set Up and Import Data
```{r eval=FALSE}
library(tibble)
library(tidyr)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(knitr)
data <- read.csv("data/heart_2020_cleaned.csv")
```
## Filter out NA Data
```{r eval=FALSE}
str(data)
summary(data)
data[rowSums((is.na(data))>0),]
na<-function(x){
  result<- which(x=='na'| x=='NA'|x=='Na')
  length(result)
}
apply(data,2,na)
```
## Change Yes/No Response to 1/0
```{r eval=FALSE}
# creating dummy variables using a function
# change Yes to 1, No to 0
convert_yes_no_to_binary <- function(data, column) {
  for (x in column) {
    data[[x]] <- ifelse(data[[x]] == 'Yes', 1, 0)
  }
  return(data)
}

dummy_variables <- c("HeartDisease", "Smoking", "AlcoholDrinking", "Stroke", "DiffWalking","Diabetic", "Asthma", "KidneyDisease", "SkinCancer")

# physical activity has a relative positive impact, so change Yes to 0, No to 1
data$PhysicalActivity <- ifelse(data$PhysicalActivity == "Yes", 0, 1)

data <- convert_yes_no_to_binary(data, dummy_variables)
head(data)
```
## Seperate Into 8 Different Group Based on Age and Sex
```{r eval=FALSE}
# check unique values in AgeCategory and Sex
unique(data$AgeCategory)
length(unique(data$AgeCategory))
unique(data$Sex)

# update AgeCategory variable
data <- data %>%
  mutate(AgeCategory = case_when(
    AgeCategory %in% c("18-24", "25-29", "30-34", "35-39") ~ "under 40",
    AgeCategory %in% c("40-44", "45-49", "50-54") ~ "40-54",
    AgeCategory %in% c("55-59", "60-64", "65-69") ~ "55-69",
    AgeCategory %in% c("70-74", "75-79", "80 or older") ~ "over 70",
    TRUE ~ "Unknown"))

# create Combined_AgeSexCategory variable using AgeCategory and Sex
data <- data %>%
  mutate(Combined_AgeSexCategory = case_when(
    AgeCategory == "under 40" & Sex == "Female" ~ "Female under 40",
    AgeCategory == "under 40" & Sex == "Male" ~ "Male under 40",
    AgeCategory == "40-54" & Sex == "Female" ~ "Female 40-54",
    AgeCategory == "40-54" & Sex == "Male" ~ "Male 40-54",
    AgeCategory == "55-69" & Sex == "Female" ~ "Female 55-69",
    AgeCategory == "55-69" & Sex == "Male" ~ "Male 55-69",
    AgeCategory == "over 70" & Sex == "Female" ~ "Female over 70",
    AgeCategory == "over 70" & Sex == "Male" ~ "Male over 70",
    TRUE ~ "Unknown"))

head(data)
```
## Create 8 Group Tables
```{r eval=FALSE}
female_under_40 <- data %>%
  filter(Combined_AgeSexCategory == "Female under 40")

male_under_40 <- data %>%
  filter(Combined_AgeSexCategory == "Male under 40")

female_40_54 <- data %>%
  filter(Combined_AgeSexCategory == "Female 40-54")

male_40_54  <- data %>%
  filter(Combined_AgeSexCategory == "Male 40-54")

female_55_69 <- data %>%
  filter(Combined_AgeSexCategory == "Female 55-69")

male_55_69 <- data %>%
  filter(Combined_AgeSexCategory == "Male 55-69")

female_over_70 <- data %>%
  filter(Combined_AgeSexCategory == "Female over 70")

male_over_70 <- data %>%
  filter(Combined_AgeSexCategory == "Male over 70")
```
## In Age Category & Sex aspect:
### Female Behaviors
```{r eval=FALSE}
# combine the data frames vertically
female_combined <- rbind(female_under_40, female_40_54, female_55_69, female_over_70)

# calculate proportions of "Yes"
props <- female_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Smoking = mean(Smoking),
    AlcoholDrinking = mean(AlcoholDrinking),
    PhysicalActivity = mean(PhysicalActivity)
  ) %>%
  gather(key = "Behavior", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70"))

# create the bar plot
ggplot(props, aes(x = Behavior, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#FEE5D9", "#FCBBA1", "#FC9272", "#DE2D26"), name = "Age/Sex Category") +
  labs(x = "Behavior", y = "Proportion of Having the Behavior", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Female Participants Who Have Each Behavior")
```
### Male Behaviors
```{r eval=FALSE}
# combine the data frames vertically
male_combined <- rbind(male_under_40, male_40_54, male_55_69, male_over_70)

# calculate proportions of "Yes"
props <- male_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Smoking = mean(Smoking),
    AlcoholDrinking = mean(AlcoholDrinking),
    PhysicalActivity = mean(PhysicalActivity)
  ) %>%
  gather(key = "Behavior", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Male under 40", "Male 40-54", "Male 55-69", "Male over 70"))

# create the bar plot
ggplot(props, aes(x = Behavior, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#EFF3FF", "#BDD7E7", "#6BAED6", "#3182BD"), name = "Age/Sex Category") +
  labs(x = "Behavior", y = "Proportion of Having the Behavior", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Male Participants Who Have Each Behavior")
```
### Female Disease
```{r eval=FALSE}
# calculate proportions of "Yes"
props <- female_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Stroke = mean(Stroke),
    DiffWalking = mean(DiffWalking),
    Diabetic = mean(Diabetic),
    Asthma = mean(Asthma),
    KidneyDisease = mean(KidneyDisease),
    SkinCancer = mean(SkinCancer)
  ) %>%
  gather(key = "Disease", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70"))

# create the bar plot
ggplot(props, aes(x = Disease, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#FEE5D9", "#FCBBA1", "#FC9272", "#DE2D26"), name = "Age/Sex Category") +
  labs(x = "Disease", y = "Proportion of Having the Disease", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Female Participants Who Responded 'Yes' to Each Disease")
```
### Male Disease
```{r eval=FALSE}
# calculate proportions of "Yes"
props <- male_combined %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    Stroke = mean(Stroke),
    DiffWalking = mean(DiffWalking),
    Diabetic = mean(Diabetic),
    Asthma = mean(Asthma),
    KidneyDisease = mean(KidneyDisease),
    SkinCancer = mean(SkinCancer)
  ) %>%
  gather(key = "Disease", value = "Prop_Yes", -Combined_AgeSexCategory)

props$Category <- factor(props$Combined_AgeSexCategory, levels = c("Male under 40", "Male 40-54", "Male 55-69", "Male over 70"))

# create the bar plot
ggplot(props, aes(x = Disease, y = Prop_Yes, fill = Category)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("#EFF3FF", "#BDD7E7", "#6BAED6", "#3182BD"), name = "Age/Sex Category") +
  labs(x = "Disease", y = "Proportion of Having this Disease", fill = "Age/Sex Category") +
  theme_bw() +
  geom_text(aes(label = round(Prop_Yes, 2)), position = position_dodge(width = 1), vjust = -0.5) + 
  ggtitle("Proportion of Male Participants Who Responded 'Yes' to Each Disease")
```
### Evaluate Each Group and Collect Score
```{r eval=FALSE}
score <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Behavior_Score = c(15, 13, 12, 12, 14, 13, 12, 11),
  Disease_Score = c(33, 30, 26, 21, 34, 32, 26, 19),
  Total_Score = c(48, 43, 38, 33, 48, 45, 38, 30)
)

score

score_sep <- score %>%
  separate(Group, c("Gender", "Age"), sep = " ", remove = FALSE)
score_sep$Age <- recode(score_sep$Age, "under" = "under 40")
score_sep$Age <- recode(score_sep$Age, "over" = "over 70")
score_sep$Age <- factor(score_sep$Age, levels = c("under 40", "40-54", "55-69", "over 70"))

line_data <- score_sep %>%
  select(-Group) %>%
  tidyr::gather(key = "Score_Type", value = "Score", -Gender, -Age)

# create the line plot
ggplot(line_data, aes(x = Age, y = Score, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Score") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  scale_color_manual(values = c("Female" = "magenta", "Male" = "blue"), name = "Gender") +
  facet_wrap(~Score_Type, ncol = 1) +
  ggtitle("Evaluation Score by Age Group")
```
### Physical Health Rating
```{r eval=FALSE}
# calculate means of PhysicalHealth by age/sex category
means <- data %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    PhysicalHealth_mean = mean(PhysicalHealth)
  )

age_groups <- c("under 40", "40-54", "55-69", "over 70")

colors <- c("Female" = "#DE2D26", "Male" = "#3182BD")

# create the line plot
mean_data <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Mean = means$PhysicalHealth_mean[c(4, 1:3, 8, 5:7)],
  Gender = c("Female", "Female", "Female", "Female", "Male", "Male", "Male", "Male"),
  Age = age_groups
)

ggplot(mean_data, aes(x = Age, y = Mean, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Average Physical Health Rating") +
  theme_bw() +
  scale_color_manual(values = colors, name = "Gender") +
  scale_x_discrete(limits = age_groups) +
  ggtitle("Average of Physical Health Rating for Female and Male on Age Range")
```
### Mental Health Rating
```{r eval=FALSE}
# calculate means of MentalHealth by age/sex category
means <- data %>%
  group_by(Combined_AgeSexCategory) %>%
  summarize(
    MentalHealth_mean = mean(MentalHealth)
  )

age_groups <- c("under 40", "40-54", "55-69", "over 70")

colors <- c("Female" = "#DE2D26", "Male" = "#3182BD")

# create the line plot
mean_data <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Mean = means$MentalHealth_mean[c(4, 1:3, 8, 5:7)],
  Gender = c("Female", "Female", "Female", "Female", "Male", "Male", "Male", "Male"),
  Age = age_groups
)

ggplot(mean_data, aes(x = Age, y = Mean, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Average Mental Health Rating") +
  theme_bw() +
  scale_color_manual(values = colors, name = "Gender") +
  scale_x_discrete(limits = age_groups) +
  ggtitle("Average of Mental Health Rating for Female and Male on Age Range")
```
### Different Group & Sleeptime
```{r eval=FALSE}
data$Combined_AgeSexCategory <- factor(data$Combined_AgeSexCategory, levels = c("Female under 40", "Male under 40", "Female 40-54", "Male 40-54", "Female 55-69", "Male 55-69", "Female over 70", "Male over 70"))

# create the box plot
ggplot(data, aes(x = Combined_AgeSexCategory, y = SleepTime, fill = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", color = "black", fill = "white", position = position_dodge(0.75)) +
  labs(x = "Age Sex Category", y = "SleepTime", fill = "Sex") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  coord_cartesian(expand = FALSE) +
  ggtitle("Boxplot of Sleep Time by Age Sex Category")
```
### Different Groups & BMI
```{r eval=FALSE}
data$Combined_AgeSexCategory <- factor(data$Combined_AgeSexCategory, levels = c("Female under 40", "Male under 40", "Female 40-54", "Male 40-54", "Female 55-69", "Male 55-69", "Female over 70", "Male over 70"))

# create the box plot
ggplot(data, aes(x = Combined_AgeSexCategory, y = BMI, fill = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", color = "black", fill = "white", position = position_dodge(0.75)) +
  labs(x = "Age Sex Category", y = "BMI", fill = "Sex") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  coord_cartesian(expand = FALSE) +
  ggtitle("BMI by Age Sex Category")
```
### Proportion of Getting Heart Disease in Different Age Group
```{r eval=FALSE}
# calculate the proportion of people with heart disease in each group
female_under_40_prop <- sum(female_under_40$HeartDisease == 1) / nrow(female_under_40)
female_40_54_prop <- sum(female_40_54$HeartDisease == 1) / nrow(female_40_54)
female_55_69_prop <- sum(female_55_69$HeartDisease == 1) / nrow(female_55_69)
female_over70_prop <- sum(female_over_70$HeartDisease == 1) / nrow(female_over_70)
male_under_40_prop <- sum(male_under_40$HeartDisease == 1) / nrow(male_under_40)
male_40_54_prop <- sum(male_40_54$HeartDisease == 1) / nrow(male_40_54)
male_55_69_prop <- sum(male_55_69$HeartDisease == 1) / nrow(male_55_69)
male_over70_prop <- sum(male_over_70$HeartDisease == 1) / nrow(male_over_70)

# combine the proportion data
prop_data <- data.frame(
  Group = c("Female under 40", "Female 40-54", "Female 55-69", "Female over 70",
            "Male under 40", "Male 40-54", "Male 55-69", "Male over 70"),
  Proportion = c(female_under_40_prop, female_40_54_prop, female_55_69_prop, female_over70_prop,
                 male_under_40_prop, male_40_54_prop, male_55_69_prop, male_over70_prop),
  Gender = c("Female", "Female", "Female", "Female", "Male", "Male", "Male", "Male"),
  Age = c("under 40", "40-54", "55-69", "over 70", "under 40", "40-54", "55-69", "over 70")
)

# create the line plot
ggplot(prop_data, aes(x = Age, y = Proportion, group = Gender, color = Gender)) +
  geom_line(size = 0.5) +
  geom_point(size = 1.5) +
  labs(x = "Age Group", y = "Proportion of people with heart disease", title= "Proportion of Getting Heart Disease in Different Age Group and Gender") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  scale_color_manual(values = c("Female" = "magenta", "Male" = "blue"), name = "Gender") +
  scale_x_discrete(limits = c("under 40", "40-54", "55-69", "over 70"))
```
## In BMI aspect:
### BMI and Behavior
```{r eval=FALSE}
behavior_data <- data %>%
  mutate(BMI_Group = cut(BMI, c(0, 10, 18, 24, 29, 34, 39, 45, 50, 60, Inf))) %>%
  group_by(BMI_Group) %>%
  summarize(HSmoking = mean(Smoking),
            AlcoholDrinking = mean(AlcoholDrinking),
            PhysicalActivity = mean(PhysicalActivity)) %>%
  pivot_longer(-BMI_Group, names_to = "Behavior", values_to = "Proportion")

# create a line plot with different colored lines for each behavior
ggplot(behavior_data, aes(x = BMI_Group, y = Proportion, color = Behavior, group = Behavior)) +
  geom_line() +
  labs(x = "BMI", y = "Proportion of Behavior", color = "Behavior") +
  theme_bw() +
  ggtitle("Proportion of Each Behavior by BMI")
```
### BMI and Other Diseases
```{r eval=FALSE}
# create a data frame with BMI groups and proportion of people getting each disease
disease_data <- data %>%
  mutate(BMI_Group = cut(BMI, c(0, 10, 18, 24, 29, 34, 39, 45, 50, 60, Inf))) %>%
  group_by(BMI_Group) %>%
  summarize(HeartDisease = mean(HeartDisease),
            Stroke = mean(Stroke),
            DiffWalking = mean(DiffWalking),
            Diabetic = mean(Diabetic),
            Asthma = mean(Asthma),
            KidneyDisease = mean(KidneyDisease),
            SkinCancer = mean(SkinCancer)) %>%
  pivot_longer(-BMI_Group, names_to = "Disease", values_to = "Proportion")

# create a line plot with different colored lines for each disease
ggplot(disease_data, aes(x = BMI_Group, y = Proportion, color = Disease, group = Disease)) +
  geom_line() +
  labs(x = "BMI", y = "Proportion of Disease", color = "Disease") +
  theme_bw() +
  ggtitle("Proportion of Getting Each Disease by BMI")
```
### BMI and HeartDisease
```{r eval=FALSE}
# create a new column with BMI group
data <- data %>%
  mutate(BMI_group = cut(BMI, breaks = c(10, 18, 24, 29, 34, 39, 45, 50, 60, Inf)))

# calculate the proportion of people getting HeartDisease
df <- data %>%
  group_by(BMI_group) %>%
  summarize(prop = mean(HeartDisease))

# create the dot plot
ggplot(df, aes(x = BMI_group, y = prop)) +
  geom_line() +
  geom_point() +
  labs(x = "BMI", y = "Proportion of people with HeartDisease") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Proportion of People with Heart Disease by BMI")
```
## In Race aspects:
### Proportion of Getting Heart Disease for Different Races and Age Groups
```{r eval=FALSE}
data_HD <- data %>% 
  mutate(prop_heartdisease = ifelse(HeartDisease == 1, 1, 0)) %>%
  group_by(AgeCategory, Race) %>%
  summarize(prop_heartdisease = mean(prop_heartdisease))

age_levels <- c("under 40", "40-54", "55-69", "over 70")

# create the co-variation plot
data_HD %>% 
  transmute(AgeCategory = factor(AgeCategory, levels = age_levels), Race, 
            prop_heartdisease, fill = prop_heartdisease) %>%
  ggplot(aes(x = AgeCategory, y = Race, fill = fill)) +
  geom_tile() +
  scale_fill_gradient(low = "#FCBBA1", high = "red") +
  labs(x = "Age Category", y = "Race", fill = "Proportion of Heart Disease") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Proportion of People with Heart Disease by Race and Age Category")
```
### Race and Heart Disease
```{r eval=FALSE}
data %>%
  group_by(Race) %>%
  summarise(prop_heartdisease = mean(HeartDisease)) %>%
  ggplot(aes(x = Race, y = prop_heartdisease, color = Race)) +
  geom_point(size = 3) +
  labs(x = "Race", y = "Proportion of Heart Disease") +
  theme_bw() +theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Proportion of People with Heart Disease by Race")
```
## Regression:
### Logistic Regression Model
```{r eval=FALSE}
# fitting the full model
reg_formula <- HeartDisease ~ BMI + SleepTime + Smoking + AlcoholDrinking + PhysicalActivity + DiffWalking + Diabetic + Asthma + KidneyDisease + SkinCancer + Stroke + PhysicalHealth + MentalHealth + AgeCategory + Sex + Race + GenHealth

full_model <- glm(reg_formula, data = data, family = binomial(link = "logit"))
summary(full_model)
```
### Variable Selection
```{r eval=FALSE}
# variable selection using stepwise AIC
library(MASS)
stepwise_model <- stepAIC(full_model, direction = "both")
summary(stepwise_model)
```
### Check for multicollinearity using Variance Inflation Factor (VIF)
```{r eval=FALSE}
library(car)
vif_values <- vif(stepwise_model)
vif_values
```
### Model interpretation
```{r eval=FALSE}
# check odds ratios for each predictor
exp(coef(stepwise_model))
```
### Perform the Likelihood Ratio Test 
```{r eval=FALSE}
# compare the stepwise model with the full model using ANOVA
anova(stepwise_model, full_model, test = "Chisq")
```